National Longitudinal Evidence for Growth in Subjective Wellbeing from Spiritual Beliefs

### 2020-11-26 Manuscript_Bayesian.Rmd
improved

Statistical analysis was performed using R version 4.0.2 R version 3.6.1 (2020-06-22) 2019-07-05). We analyze data from participants who responded to the NZAVS survey at least three times between 2010 (Time 2) and 2018 (Time 10), resulting in a sample of  21,705 (N = 20,979 with complete responses).  To model growth in subjective well-being we use generalized linear mixed models. In our models, we account for the non-linear effects of time within participants by including the interaction of years (a continuous variable) with Beliefs (a four-level categorical indicator). The baseline for this indicator is Skepticism about a Spirit/Life Forces and Skepticism about a God. To handle the dependencies and heterogeneity introduced from the repeated measures, we include individual ID as an effect modelled as random.

@@

In a supplemental analysis we include standard demographic indicators used in New Zealand Attitudes and Values Research: [NZAVS modelling standards](https://cdn.auckland.ac.nz/assets/psych/about/our-research/nzavs/NZAVSTechnicalDocuments/NZAVS-Technical-Documents-e11-Default-Statistical-Standards.pdf). We centered and scaled the education, deprivation, and political conservativism. We centred age was centred at and put into decade-long intervals. The model equation(s) are described in Equation 2.

Because missing values can result in biased estimates [@Blackwell2017-oq], we multiply imputed 10 datasets using the AmeliaII package in R[@Honaker2011-yu].We conducted a parallel analysis averaging over the multiply imputed datasets, without any substantive difference to inferences. Additionally the results were robust to filtering of participants of N >3 to N = 9 waves. Additionally, the findings reported below were robust to modeling only the effects of years and beliefs, without adjusting for the standard NZAVS demographic covariates. The supplement describing these analysis is located at: https://github.com/jacopastorius/nzavs_spiritbeliefWellbeing.

@@

All code for this analysis and a full report of the missing data imputation can be found at https://github.com/jacopastorius/nzavs_spiritbeliefWellbeing.

@@

```{r fig.cap = "Combination Coefficient Plots for personal well-being and life satisfaction longitudinal growth models"}
knitr::include_graphics("/Users/joseph/Documents/GitHub/nzavs_spiritbeliefWellbeing/Figs/USE_2_coefficientplots-1.png")
```

@@

```{r fig.cap = "Combination Coefficient Plots for personal well-being and life satisfaction longitudinal growth models"}
knitr::include_graphics("/Users/joseph/Documents/GitHub/nzavs_spiritbeliefWellbeing/Figs/USE_2_coefficientplots-1.png")

@@

1. At the outset of the study, in 2010, the expected subjective well-being as measured using the personal well-being Inventory was expected to be lower among those who believed in Spirit or Life Force but not a God ($\hat{y}$ =  6.78 [6.74, 6.82]) and lower among those who believed in both a Spirit/Life-Force and a God ($\hat{y}$ =  6.75 [6.72, 6.78]) in comparison both to the skeptical population who neither believed in a God nor a life force ($\hat{y}$ = 6.91 [6.86, 6.95]) and to the population that believed in a God but was skeptical about belief in a Spirit or Life Force ($\hat{y}$ 6.88 [6.81, 6.96]). (See Figure 2)  Similarly,the expected subjective well-being as measured using the life satisfaction Inventory was lowest in 2010 among those who believed in Spirit or Life Force but not a God ($\hat{y}$ = 5.07 [5.04, 5.10]). Life satisfaction was similar among those who believed in both a Spirit/Life-Force and a God  ($\hat{y}$ =  5.11 [5.08, 5.13]) in comparison both with the skeptical population who neither believed in a God nor a life force ($\hat{y}$ = 5.12 [5.08, 5.15]). Life satisfaction was highest among the population that believed in a God but was skeptical about belief in a Spirit or Life Force ($\hat{y}$ 5.19 [5.14, 5.25]) (see Figure 3)

2. However, over time, belief in a Spirit or Life Force predicts a trajectory of growth in personal well-being. In 2018, personal well-being among who believe in Spirit or Life Force but not a God had grown to ($\hat{y} = 7.13$ [7.09, 7.16], $\Delta \hat{y} = 0.35$). Among those who believed in both a Spirit/Life-Force and a God, personal well-being had grown to ($\hat{y} =  7.14$, [7.11, 7.16], $\Delta \hat{y} = 0.39$). By contrast, personal well-being among the skeptical population who neither believed in a God nor a Life Force grew less steeply across the eight year interval $\hat{y} = 7.15 [7.12, 7.19]$, $\Delta \hat{y} = 0.24$). By the same token, personal well-being among those who believe in a God but are skeptical about a Spirit/Life Force grew less steeply $\hat{y} = 7.15$, [7.09, 7.21] $\Delta \hat{y} = 0.27$). Similarly, over time, belief in a Spirit or Life Force predicts a trajectory of growth in life satisfaction. In 2018, the expected effect of spiritual beliefs on life satisfaction among those who believe in Spirit or Life Force but not a God had grown to ($\hat{y} =  5.30$ [5.28, 5.33], $\Delta \hat{y} = 0.23$). In practical terms, such growth is equivalent to the expected gain in life satisfaction from being employed and a sixth as large as the expected benefit to life satisfaction from having a romantic partner. In practical termss, the rate of increase in personal wellbeing among those with Spirit/Life force beliefs was about one and half times greater than it was among those who lacked such beliefs (See Figure 2). Among those who believed in both a Spirit/Life-Force and a God, life satisfaction had grown to ($\hat{y} = 5.33 [5.31, 5.35]$, $\Delta \hat{y} = 0.22$). By contrast, life satisfaction among the skeptical population who neither believed in a God nor a Life Force lagged behind over the eight year interval $\hat{y} =  5.24 [5.21, 5.27], \Delta \hat{y} = 0.12$). Relatedly, life satisfaction in the population that believe in God but are skeptical about a Spirit/Life Force grew less steeply $\hat{y} =  5.32 [5.27, 5.36], \Delta \hat{y} = 0.13$). In practical terms, the rate of increase in life satisfaction among those with Spirit/Life force beliefs was twice as large (See Figure 3).

@@

## Supplement

## Personal Well-Being

We fitted a linear mixed model (estimated using REML and nloptwrap optimizer) to predict personal well-being (PWI) with years (0-9), beliefs (four categories), Age.10yrs.C, deprivation (S), education (S), employed, ethnic categories, male, partner, political conservativism (S) and urban, with Id as a random effect. The model's intercept, corresponding to years = 0, beliefs = Skeptic, age = (sample mean), deprivation.S = (sample mean), education (S) =(sample mean), employed = not employed, ethnic Categories = European, male = not male, partner = no partner, political conservativism (S) = (sample mean), urban = not urban, is at 4.94 (SE = 0.02, 95% CI [4.89, 4.99], p < .001) is at 6.63 (SE = 0.03, 95% CI [6.57, 6.70], p < .001). Within this model:

  - The effect of Years is positive is not reliable (beta = 2.66e-03, SE = 3.58e-03, std. beta = 3.19e-03, p = 0.458).  We infer that personal well-being did not increase in the Skeptic population between 2010 and 2018.
  - The effect of Beliefs in *Spirit Excludes God*  is reliably negative (beta = -0.14, SE = 0.03, std. beta = -0.03, p < .001). We infer that in 2010, the population that believed in a Spirit but did not believe in a God had lower personal well-being than did Skeptics. 
  - The effect of Beliefs in *God And Spirit*  is reliably negative (beta = -0.16, SE = 0.03, std. beta = -0.02, p < .001). We infer that in 2010, the population that believed in a God and a Spirit or Life Force had lower Personal Well-being than did Skeptics. 
  - The effect of Beliefs in *God Excludes Spirit*  is not reliable (beta = 0.02, SE = 0.05, std. beta = 0.02, p = 0.669). We infer that in 2010, the population that believed in a God and doubted a Spirit or Life Force similar levels of personal well-being to Skeptics.
  - The effect of Age.10yrs.C is reliably positive  (beta = 0.08, SE = 7.38e-03, std. beta = 0.07, p < .001). Older people tend to have greater personal well-being.
  - The effect of Deprivation (S) is reliably negative (beta = -0.14, SE = 7.50e-03, std. beta = -0.08, p < .001). Deprived people tend to have lower personal well-being.
  - The effect of Education (S) is reliably positive  (beta = 0.12, SE = 8.74e-03, std. beta = 0.07, p < .001). Educated people tend to have greater personal well-being.
  - The effect of Employed is reliably positive  (beta = 0.09, SE = 0.01, std. beta = 0.05, p < .001).
  - The effect of identifying as M\={a}ori is reliably negative (beta = -0.30, SE = 0.03, std. beta = -0.18, p < .001).
  - The effect of identifying as Pacific is reliably negative (beta = -0.29, SE = 0.05, std. beta = -0.17, p < .001). 
  - The effect of identifying as Asian is not reliable (beta = -0.08, SE = 0.05, std. beta = -0.05, p = 0.089).
  - The effect of identifying as male is reliably negative (beta = -0.11, SE = 0.02, std. beta = -0.06, p < .001). 
  - The effect of having a partner is reliably positive (beta = 0.61, SE = 0.02, std. beta = 0.37, p < .001).
  - The effect of political conservativism (S) is reliably positive  (beta = 0.02, SE = 5.67e-03, std. beta = 0.01, p < .001). 
  - The effect of living in an Urban area is reliably negative (beta = -0.02, SE = 0.01, std. beta = -0.01, p < .05). T

##  Personal well-being: years X beliefs

```{r fig.cap = "Combination Coefficient plots for personal well-being and life satisfaction longitudinal growth models"}
knitr::include_graphics("Figs/USEPWI_expected-1.png")
```
Focusing on the interaction of time with belief-type on personal well-being we observe:

  - The effect of years $\times$ belief in a *Spirit Excludes God*  is reliably positive  (beta = 0.02, SE = 4.86e-03, std. beta = 0.02, p < .01). We infer that, unlike the skeptic population, people with beliefs in a spirit or life force who do not believe in a God tend to grow in their personal well-being over time.
  - The effect of years $\times$ in *God And Spirit*  is reliably positive  (beta = 0.02, SE = 4.41e-03, std. beta = 0.03, p < .001). We infer that people who believe in a spirit or life force who also believe in a God tend to grow in their personal well-being over time
  - The effect of Years $\times$ belief in *God Excludes Spirit*  is not reliable(beta = 1.96e-03, SE = 8.29e-03, std. beta = 2.36e-03, p = 0.813).  Thus, God believers who doubt in a Spirit or Life Force start maintain higher personal well-being than do other groups, but they are not expected to grow in personal well-being over time.

## Life Satisfaction

We fitted a linear mixed model (estimated using REML and nloptwrap optimizer) to predict personal well-being (PWI) with years (0-9), beliefs (four categories), Age.10yrs.C, deprivation (S), education (S), employed, ethnic categories, male, partner, political conservativism (S) and urban, with Id as a random effect. Standardized parameters were obtained by fitting the model on a standardized version of the dataset. The model's total explanatory power is substantial (conditional R2 = 0.69). The model's intercept, corresponding to years = 0, beliefs = Skeptic, age = (sample mean), deprivation.S = (sample mean), education (S) =(sample mean), employed = not employed, ethnic Categories = European, male = not male, partner = no partner, political conservativism (S) = (sample mean), urban = not urban is at 4.94 (SE = 0.02, 95% CI [4.89, 4.99], p < .001). Within this model:

  - The effect of years is not reliable (beta = -5.50e-04, SE = 2.69e-03, std. beta = -9.32e-04, p = 0.838).  We infer that life satisfaction did not increase in the Skeptic population between 2010 and 2018.
  - The effect of beliefs in *Spirit Excludes God* is reliably negative (beta = -0.08, SE = 0.02, std. beta = 0.01, p < .001). We infer that overall the population that believes in a Spirit but does not believe in a God is expected to have a lower Life Satisfaction than skeptics. 
  - The effect of beliefs in *God And Spirit* is not reliable (beta = -0.04, SE = 0.02, std. beta = 0.04, p = 0.105). People who believe in God but doubt a Spirit/Life force are expected to have had similar levels of life satisfaction in Wave 10 as skeptics.
  - The effect of beliefs in *God Excludes Spirit* is reliably positive (beta = 0.07, SE = 0.04, std. beta = 0.05, p = 0.062).We infer that overall the population that believes in a God but does not believe in a spirit is not reliably different from  the Skeptics population in Life Satisfaction. 
  - The effect of Age.10yrs.C is reliably positive (beta = 0.09, SE = 5.26e-03, std. beta = 0.10, p < .001). Older people have greater Life Satisfaction
  - The effect of Deprivation (S) is  is reliably negative (beta = -0.07, SE = 5.50e-03, std. beta = -0.06, p < .001). People who are more deprived have lower Life Satisfaction.
  - The effect of Education (S) is reliably positive (beta = 0.06, SE = 6.33e-03, std. beta = 0.05, p < .001). Educated people have greater Life Satisfaction
  - The effect of being Employed is reliably positive (beta = 0.08, SE = 9.69e-03, std. beta = 0.07, p < .001). People who are employed tend to have greater Life Satisfaction
  - The effect of identifying as Maori is reliably negative (beta = -0.06, SE = 0.02, std. beta = -0.05, p < .01).  People who belong to the minority Maori group tend to have lower Life Satisfaction.
  - The effect of identifying as Pacific is reliably negative (beta = -0.08, SE = 0.04, std. beta = -0.07, p < .05).  People who belong to the minority Pacific peoples group tend to have lower life satisfaction.
  - The effect of identifying as Asian is not reliably different from people of European ethnicity in life satisfaction(beta = -0.04, SE = 0.03, std. beta = -0.03, p = 0.223).
  - The effect of identifying as male is reliably negative (beta = -0.19, SE = 0.01, std. beta = -0.16, p < .001).
  - The effect of having a partner is reliably positive (beta = 0.42, SE = 0.01, std. beta = 0.35, p < .001).
  - The effect of political conservativism (S) is reliably positive (beta = 0.01, SE = 4.23e-03, std. beta = 0.01, p < .01).
  - The effect of living in an urban area is reliably negative (beta = -0.02, SE = 8.85e-03, std. beta = -0.02, p < .05).

## Life Satisfaction: years X beliefs

```{r fig.cap = "Prediction Plot Life Satisfaction Spirit Belief Longitudinal Models"}
knitr::include_graphics("/Users/joseph/Documents/GitHub/nzavs_spiritbeliefWellbeing/Figs/USElifsat_expected-1.png")
```
  - The interaction of years $\times$ beliefs in *Spirit Excludes God* ludesGod is reliably positive (beta = 0.02, SE = 3.66e-03, std. beta = 0.03, p < .001). We infer that, unlike the Skeptic population, people with beliefs in a Spirit or Life Force who do not believe in a God tend to grow in their Life Satisfaction over time.
  - The interaction of years $\times$ *God And Spirit*  is reliably positive (beta = 0.01, SE = 3.32e-03, std. beta = 0.02, p < .001). We infer that people who believe in a spirit or life force who also believe in a God tend to grow in their life satisfaction over time
  - The effect of years $\times$ beliefs in *God Excludes Spirit*  is not reliable (beta = -1.59e-03, SE = 6.25e-03, std. beta = -2.70e-03, p = 0.799). Thus, God believers who doubt in a Spirit or Life Force start maintain higher Life-Satisfaction than do other groups, but they are not expected to grow in Life-Satisfaction over time.

## Imputed data analysis
To adjust for potential bias from missing responses, we conducted a supplementary analysis in which we imputed missing values over the the longitudinal sample. This analysis is describe in Supplement 1. Averaging over the missing responses does not change inference, however we observe growth across all four groups in both personal well-being and life satisfaction. Additionally we observe substantially stronger effects for the slope of Spirit Beliefs on both personal well-being and life-satisfaction. Figures 3 and Figures 4 describe the predicted slopes of change. The inference from this analysis remains the same as from the analysis in which we delete missing values row-wise, by panel wave. In all analyses, those who believe in a spirit or life-force experience greater growth in both personal well-being and life satisfaction over time.

Figure 3 about here:

```{r fig.cap = "Imputed data analysis: coefficient plot"}
knitr::include_graphics("Figs/coefficient_plot_both_imputed-1.png")
```

Figure 4 about here:

```{r fig.cap = "Imputed data analysis: predicted personal wellbeing by years X beliefs"}
knitr::include_graphics("/Users/joseph/Documents/GitHub/nzavs_spiritbeliefWellbeing/Figs/predicted_PWI_impute-1.png")
```

Figure 5 about here:

```{r fig.cap = "Imputed data analysis: predicted personal wellbeing by years X beliefs"}
knitr::include_graphics("/Users/joseph/Documents/GitHub/nzavs_spiritbeliefWellbeing/Figs/predicted_lifesat_imputed-1.png")
```

\newpage
# Supplement Model Equations

## Priors for the Personal Wellbeing model

@@

\newpage
## Priors for the Life Satisfaction model

Model priors for Life Satisfaction Model with including group-level ID intercept and belief slopes are as follows:
$$
\begin{aligned}
\operatorname{Life Satisfaction}_{ij} \sim \text{Normal}(\mu_{ij},\sigma^2) \nonumber\\
\mu_{ij}\sim \boldsymbol{\alpha +\beta} \nonumber\\
\boldsymbol{\alpha} \sim \alpha_{0} +\alpha_j \nonumber\\
\alpha_{0} \sim \text{StudentT}(3,5.5, 2.5)\nonumber\\
\alpha_j \sim \text{StudentT}+(3,0,10)\nonumber\\
\boldsymbol{\beta}\sim \boldsymbol{b} + \beta_j \nonumber\\
\boldsymbol{b}\sim\text{Normal}(0,2.5) \nonumber\\
\begin{bmatrix}
\alpha_j \\
\beta_j
\end{bmatrix}
\sim 
\text{MVNormal}
\begin{pmatrix}
\begin{bmatrix}
0\\
0
\end{bmatrix}
,\boldsymbol{S} \nonumber
\end{pmatrix}\\
\boldsymbol{S} = 
\begin{pmatrix}
\sigma_{\alpha_j} & 0 \\
0 & \sigma_{\beta_j} \nonumber
\end{pmatrix} \boldsymbol{R} \begin{pmatrix}
\sigma_{\alpha_j} & 0 \\
0 & \sigma_{\beta_j} \nonumber
\end{pmatrix}\\
\sigma_{\alpha_j} \sim 
\text{HalfCauchy}(0,1) \nonumber\\
\sigma_{\beta_j} \sim 
\text{HalfCauchy}(0,1) \nonumber\\
\boldsymbol{R}\sim \text{LKJcorr}(1) \nonumber\\
\sigma  \sim  \operatorname{HalfStudent T}^{+}(3, 0, 2.5) 
\end{aligned}
$$
\newpage

## Deterministic model

>>>>

Statistical analysis was performed using R version 4.0.2 R version 3.6.1 (2020-06-22) 2019-07-05). We analyze data from participants who responded to the NZAVS survey at least three times between 2010 (Time 2) and 2018 (Time 10), resulting in a sample of  21,705 (N = 20,979 with complete responses).  To model growth in subjective well-being we use generalized linear mixed models. In our models, we account for the non-linear effects of time within participants by including the interaction of years (a continuous variable) with Beliefs (a four-level categorical indicator). The baseline for this indicator is Skepticism about a Spirit/Life Forces and Skepticism about a God. To handle the dependencies introduced from the repeated measures, we include individual ID as an effect modeled as random, and estimated a random effect for the slope for years. The model equations are given below and priors are given below.

@@

Because missing values can result in biased estimates [@Blackwell2017-oq], we multiply imputed 10 datasets using the AmeliaII package in R[@Honaker2011-yu].We conducted a parallel analysis averaging over the multiply imputed datasets, without any substantive difference to inferences. Additionally the results were robust to filtering of participants of N > 3 to N = 8 waves. In a supplemental analysis we include standard demographic indicators used in New Zealand Attitudes and Values Research: [NZAVS modelling standards](https://cdn.auckland.ac.nz/assets/psych/about/our-research/nzavs/NZAVSTechnicalDocuments/NZAVS-Technical-Documents-e11-Default-Statistical-Standards.pdf). We centered and scaled the education, deprivation, and political conservativism. We centred age was centred at and put into decade-long intervals. There are increasing worries

We provide the analyses of these additional analyses in the supplement.  The supplement describing these analysis is located at: https://github.com/jacopastorius/nzavs_spiritbeliefWellbeing.

@@

All code for this analysis and a full report of the missing data imputation can be found at [https://github.com/jacopastorius/nzavs_spiritbeliefWellbeing](https://github.com/jacopastorius/nzavs_spiritbeliefWellbeing).

@@

```{r fig.cap = "Combination Coefficient Plots for personal well-being and life satisfaction longitudinal growth models"}
knitr::include_graphics("/Users/joseph/Documents/GitHub/nzavs_spiritbeliefWellbeing/Bayesian_pwi_predicted.png")
```

@@

\begin{table}
\begin{center}
\scalebox{0.5}{
\begin{tabular}{l c c}
\toprule
 & PWI & Life Sat \\
\midrule
Intercept                           & $\mathbf{7.04}$   & $\mathbf{5.19}$  \\
                                    & $ [ 7.02;  7.07]$ & $ [ 5.17; 5.21]$ \\
yearsC                              & $\mathbf{0.02}$   & $\mathbf{0.01}$  \\
                                    & $ [ 0.02;  0.03]$ & $ [ 0.01; 0.02]$ \\
Beliefs\_SpiritExcludesGod\_        & $\mathbf{-0.06}$  & $0.02$           \\
                                    & $ [-0.09; -0.03]$ & $ [-0.00; 0.04]$ \\
BeliefsGodAndSpirit                 & $\mathbf{-0.06}$  & $\mathbf{0.05}$  \\
                                    & $ [-0.10; -0.03]$ & $ [ 0.03; 0.07]$ \\
BeliefsGodExcludesSpirit            & $-0.00$           & $\mathbf{0.08}$  \\
                                    & $ [-0.04;  0.04]$ & $ [ 0.04; 0.11]$ \\
yearsC:Beliefs\_SpiritExcludesGod\_ & $\mathbf{0.01}$   & $\mathbf{0.01}$  \\
                                    & $ [ 0.00;  0.02]$ & $ [ 0.01; 0.02]$ \\
yearsC:BeliefsGodAndSpirit          & $\mathbf{0.01}$   & $\mathbf{0.01}$  \\
                                    & $ [ 0.01;  0.02]$ & $ [ 0.00; 0.02]$ \\
yearsC:BeliefsGodExcludesSpirit     & $0.00$            & $-0.00$          \\
                                    & $ [-0.01;  0.02]$ & $ [-0.01; 0.01]$ \\
\midrule
SD: Idsd(Intercept)                 & $1.44$            & $0.99$           \\
SD: Idsd(yearsC)                    & $0.13$            & $0.10$           \\
SD: Idcor(Intercept,yearsC)         & $-0.02$           & $0.03$           \\
\bottomrule
\end{tabular}
}
\caption{}
\label{tab:REGRESS}
\end{center}
\end{table}

```{r fig.cap = "Combination Coefficient Plots for personal well-being and life satisfaction longitudinal growth models"}
knitr::include_graphics("/Users/joseph/Documents/GitHub/nzavs_spiritbeliefWellbeing/Bayesian_pwi_predicted.png")
```

```{r fig.cap = "Combination Coefficient Plots for personal well-being and life satisfaction longitudinal growth models"}
knitr::include_graphics("/Users/joseph/Documents/GitHub/nzavs_spiritbeliefWellbeing/Bayesian_lifesat_predicted.png")
```

@@

1. At the outset of the study, in 2010, the expected subjective well-being as measured using the personal well-being Inventory was expected to be lower among those who believed in Spirit or Life Force but not a God ($\hat{y} =$  6.80 [6.76, 6.85]) and lower among those who believed in both a Spirit/Life-Force and a God ($\hat{y}$ =  6.78 [6.74, 6.81]) in comparison both to the skeptical population who neither believed in a God nor a life force ($\hat{y}$ = 6.92 [6.87, 6.96]) and to the population that believed in a God but was skeptical about belief in a Spirit or Life Force ($\hat{y}$ 6.90 [6.83, 6.98]). (See Figure 2)  Similarly,the expected subjective well-being as measured using the life satisfaction Inventory was lowest in 2010 among those who believed in Spirit or Life Force but not a God ($\hat{y}$ = 5.07 [5.04, 5.10]). Life satisfaction was similar among those who believed in both a Spirit/Life-Force and a God  ($\hat{y}$ =  5.11 [5.08, 5.13]) in comparison both with the skeptical population who neither believed in a God nor a life force ($\hat{y}$ = 5.12 [5.08, 5.15]). Life satisfaction was highest among the population that believed in a God but was skeptical about belief in a Spirit or Life Force ($\hat{y}$ 5.19 [5.14, 5.25]) (see Figure 3)

2. However, over time, belief in a Spirit or Life Force predicts a trajectory of growth in personal well-being. In 2018, personal well-being among who believe in Spirit or Life Force but not a God had grown to ($\hat{y} =  7.12$ [7.09, 7.16], $\Delta \hat{y} = 0.32$). Among those who believed in both a Spirit/Life-Force and a God, personal well-being had grown to ($\hat{y} =  7.13$ [7.10, 7.16], $\Delta \hat{y} = 0.35$). By contrast, personal well-being among the skeptical population who neither believed in a God nor a Life Force grew less steeply across the eight year interval $\hat{y} = 7.14$ [7.11, 7.18], $\Delta \hat{y} = 0.22$). By the same token, personal well-being among those who believe in a God but are skeptical about a Spirit/Life Force grew less steeply $\hat{y} = 7.15$, [7.09, 7.21] $\Delta \hat{y} = 0.25$). Similarly, over time, belief in a Spirit or Life Force predicts a trajectory of growth in life satisfaction. In 2018, the expected effect of spiritual beliefs on life satisfaction among those who believe in Spirit or Life Force but not a God had grown to ($\hat{y} =  5.31$ [5.28, 5.33], $\Delta \hat{y} = 0.23$). In practical terms, such growth is equivalent to the expected gain in life satisfaction from being employed and a sixth as large as the expected benefit to life satisfaction from having a romantic partner. In practical terms, the rate of increase in personal wellbeing among those with Spirit/Life force beliefs was about one and half times greater than it was among those who lacked such beliefs (See Figure 2). Among those who believed in both a Spirit/Life-Force and a God, life satisfaction had grown to ($\hat{y} = 5.33 [5.31, 5.35]$, $\Delta \hat{y} = 0.24$). By contrast, life satisfaction among the skeptical population who neither believed in a God nor a Life Force lagged behind over the eight year interval $\hat{y} =  5.24 [5.21, 5.27], \Delta \hat{y} = 0.13$). Relatedly, life satisfaction in the population that believe in God but are skeptical about a Spirit/Life Force grew less steeply $\hat{y} =  5.31 [5.26, 5.36], \Delta \hat{y} = 0.11$). In practical terms, the rate of increase in life satisfaction among those with Spirit/Life force beliefs was twice as large (See Figure 3).

@@

# Appendix

\newpage
## Model Equations

### Priors for the Personal Wellbeing model

@@

### Priors for the Life Satisfaction model

Model priors for Life Satisfaction Model with including group-level ID intercept and belief slopes are as follows:

$$
\begin{aligned}
\operatorname{Life Satisfaction}_{ij} \sim \text{Normal}(\mu_{ij},\sigma^2) \nonumber\\
\mu_{ij}\sim \boldsymbol{\alpha +\beta} \nonumber\\
\boldsymbol{\alpha} \sim \alpha_{0} +\alpha_j \nonumber\\
\alpha_{0} \sim \text{StudentT}(3,5.5,2.5)\nonumber\\
\alpha_j \sim \text{StudentT}+(3,0,2.5)\nonumber\\
\boldsymbol{\beta}\sim \boldsymbol{b} + \beta_j \nonumber\\
\boldsymbol{b}\sim\text{Normal}(0,2.5) \nonumber\\
\begin{bmatrix}
\alpha_j \\
\beta_j
\end{bmatrix}
\sim 
\text{MVNormal}
\begin{pmatrix}
\begin{bmatrix}
0\\
0
\end{bmatrix}
,\boldsymbol{S} \nonumber
\end{pmatrix}\\
\boldsymbol{S} = 
\begin{pmatrix}
\sigma_{\alpha_j} & 0 \\
0 & \sigma_{\beta_j} \nonumber
\end{pmatrix} \boldsymbol{R} \begin{pmatrix}
\sigma_{\alpha_j} & 0 \\
0 & \sigma_{\beta_j} \nonumber
\end{pmatrix}\\
\sigma_{\alpha_j} \sim 
\text{HalfCauchy}(0,1) \nonumber\\
\sigma_{\beta_j} \sim 
\text{HalfCauchy}(0,1) \nonumber\\
\boldsymbol{R}\sim \text{LKJcorr}(1) \nonumber\\
\sigma  \sim  \operatorname{HalfStudent T}^{+}(3, 0, 2.5) 
\end{aligned}
$$
\newpage

### Deterministic model
